Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-12-16 1.Paper/Paper.Rmd
Update Paper.Rmd

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation, i.e., $\boldsymbol{X_i}^O=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction variable, $\boldsymbol{X_i^S}=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}},\boldsymbol{X_{3i}}]$. Then in case of a missing continuous variable, we calculate the latent variables  $\boldsymbol{y_{i}^*}$ and $\boldsymbol{r^*_{i}}$ as follows:
\begin{align*}
 \boldsymbol{y_{i}^*}&= \boldsymbol{\beta_i^O}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^O_{i}} \\
 \boldsymbol{r_{i}^*}&= \boldsymbol{\beta_i^S}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^S_{i}}
\end{align*}
Here we assumed that all coefficient parameters varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{hi}^O&= \beta_h^O + b_{hi}^O\\
 \beta_{hi}^S&= \beta_h^S + b_{hi}^S
\end{align*}
We fixed coefficients $\beta_h^O=\{0.3,1,1\}$  and $\beta_h^S=\{-0.8,1.3,-0.7,1.2\}$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we ensured that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in the data set, by removing the outcome values in the 20% of the clusters.

We assumed that random effects were independent within equations ($b_{h0}^O\indep b_{h1}^O \indep b_{h2}^O$ and $b_{h0}^S \indep b_{h1}^S \indep b_{h2}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b^O_{h} \\
b^S_{h}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\sigma_{bh}^2\begin{pmatrix}
1 & \rho*0.4 \\
\rho*0.4 &1 
\end{pmatrix}\right)
\]
with the parameters $\sigma_{b0}^2=\sigma_{b1}^2=\sigma_{b2}^2=0.4$.  We considered that the correlation parameter of the random effects between equations is 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_3\sim N(0,0.2)$ assuming that the intracluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter was given different values depending on the simulated missing mechanism (Table \ref{tab:scenarios}) .

@@

## Estimands
The estimands were the parameter coefficients of the outcome equation $\beta^O=\{\beta^O_0,\beta^O_1,\beta^O_2\}$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report the estimated variance of the random effects and residual errors $\sigma_{b0}^2$,$\sigma_{b1}^2$,$\sigma_{b2}^2$,$\sigma_{e}^2$.

@@

## Estimating procedures

@@

In the case of a missing bivariate variable, we also found (Figure \ref{fig:rhobin}) that the 2l.Heckman method provided the least unbiased results on the coefficient parameters. However, we observed more unbiased estimates and a larger 95th percentile amplitude in the estimates of the variance of the random effects, especially in the $\sigma_{b2}$ compared to the estimates obtained in the continuous incomplete outcome scenario.

@@

When we reduced the sample size to 50 patients per study, the bias and RMSE of the $\sigma_b2$ were drastically affected (not shown here but in the Appendix). This could be explained in part due to the scarce information on certain clusters, which affects directly the estimation of the Heckman model on those clusters.

@@

When using the Heckman model in an explicit MNAR process, i.e. when the probability of loss is associated with the value of the missing variable, we observe that under all imputation methods we obtain biased estimators, with poor coverage.  Particularly when estimating the intercept we observed more biased estimates compared to those obtained for the other estimated parameters B2 with coverage below 60%.

Our model may not be fully suitable for this type of scenario, especially if the main analysis is focused on estimating absolute prevalence estimates, as it is greatly affected by the bias of the intercept parameter ($\beta_0$) and the width of the 95th percentile CI of all coefficient parameters. Also the imputation method was affected in terms of the bias of the random-effects parameters $\sigma_{b2}$.

>>>>

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation, i.e., $\boldsymbol{X_i}^O=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction variable, $\boldsymbol{X_i^S}=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}},\boldsymbol{X_{3i}}]$. Then in case of a missing continuous variable, we calculate the latent variables  $\boldsymbol{y_{i}^*}$ and $\boldsymbol{r^*_{i}}$ as follows:
\begin{align*}
 \boldsymbol{y_{i}^*}&= \boldsymbol{\beta_i^O}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^O_{i}} \\
 \boldsymbol{r_{i}^*}&= \boldsymbol{\beta_i^S}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^S_{i}}
\end{align*}
Here we assumed that all coefficient parameters varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{i}^O&= \beta^O + b_{i}^O\\
 \beta_{i}^S&= \beta^S + b_{i}^S
\end{align*}
We fixed coefficients $\beta^O=\{0.3,1,1\}$  and $\beta^S=\{-0.8,1.3,-0.7,1.2\}$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we ensured that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in the data set, by removing the outcome values in the 20% of the clusters.

We assumed that random effects were independent within equations ($b_{0}^O\indep b_{1}^O \indep b_{2}^O$ and $b_{0}^S \indep b_{1}^S \indep b_{2}^S\indep b_{3}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b^O \\
b^S
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},0.4^2\begin{pmatrix}
1 & \rho*0.4 \\
\rho*0.4 &1 
\end{pmatrix}\right)
\]
for the coefficients of intercept,$X_1$,$X_2$ and $X_3$.  We considered that the correlation parameter of the random effects between equations is 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_3\sim N(0,0.2^2)$ assuming that the intracluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter was given different values depending on the simulated missing mechanism (Table \ref{tab:scenarios}) .

@@

## Estimands
The estimands were the parameter coefficients of the outcome equation $\beta^O=\{\beta^O_0,\beta^O_1,\beta^O_2\}$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report the estimated standard deviation from the covariance matrix of the random effects, i.e.,  $\sqrt{\psi_{00}}$,$\sqrt{\psi_{11}}$,$\sqrt{\psi_{22}}$.

#### Estimating procedures:

@@

@@

In the case of a missing bivariate variable, we also found (Figure \ref{fig:rhobin}) that the 2l.Heckman method provided the least unbiased results on the coefficient parameters. However, we observed more unbiased estimates and a larger 95th percentile amplitude in the estimates of the standard deviations of the random effects, especially in the $\sqrt{\psi_{22}}$ compared to the estimates obtained in the continuous incomplete outcome scenario.

@@

When we reduced the sample size to 50 patients per study, the bias and RMSE of the $\sqrt{\psi_{22}}$ were drastically affected (not shown here but in the Appendix). This could be explained in part due to the scarce information on certain clusters, which affects directly the estimation of the Heckman model on those clusters.

@@

When using the Heckman model in an explicit MNAR process, i.e. when the probability of loss is associated with the value of the missing variable, we observe that under all imputation methods we obtain biased estimators, with poor coverage.  Particularly when estimating the intercept we observed more biased estimates compared to those obtained for the other estimated parameters $\beta_2$ with coverage below 60%.

Our model may not be fully suitable for this type of scenario, especially if the main analysis is focused on estimating absolute prevalence estimates, as it is greatly affected by the bias of the intercept parameter ($\beta_0$) and the width of the 95th percentile CI of all coefficient parameters. Also the imputation method was affected in terms of the bias of the standard deviation of the random-effects parameters $\sqrt{\psi_{22}}$.
